---
title: "TEMPLATE - colors"
author: "Your Name"
date: 2026-02-04
toc: true
---

::: {.callout-note}
This template follows [lecture 5.2 slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture5.2-colors-slides.html#/title-slide){target="_blank"}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Setup

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(palmerpenguins)
library(tidyverse)

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                              create base plots                           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# requires categorical color scale ----
cat_color_plot <- ggplot(na.omit(penguins), 
                         aes(x = bill_length_mm, y = bill_depth_mm, 
                             color = species, shape = species)) +
  geom_point(size = 4, alpha = 0.8)

cat_color_plot

# requires continuous color scale ----
con_color_plot <- ggplot(na.omit(penguins), 
                         aes(x = bill_length_mm, y = bill_depth_mm, 
                             color = body_mass_g)) +
  geom_point(size = 4, alpha = 0.8) 

con_color_plot
```

## Colors for inclusive & accessible design

### Viridis scales

- Check out the [documentation](https://ggplot2.tidyverse.org/reference/scale_viridis.html) for additional viridis palette options. 

```{r}
#| eval: true
#| echo: true
#| warning: false

# discrete viridis scales ----
cat_color_plot 

# continuous color scales ----
con_color_plot
```

### RColorBrewer scales

- Exploring available palettes and generating HEX codes (or check out the [web-based interface](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)):

```{r}
#| eval: true
#| echo: true
# display only colorblind-friendly RColorBrewer palettes ----
RColorBrewer::display.brewer.all()

# preview palette with your number of desired colors ----
RColorBrewer::display.brewer.pal()

# print the HEX codes of your palette ----
RColorBrewer::brewer.pal()

# or save hex codes as a vector ----
```

- Using RColorBrewer palettes

```{r}
#| eval: true
#| echo: true
#| warning: false

# for qualitative palettes ----
cat_color_plot 

# for unclassed continuous color scales ----
con_color_plot

# for classed continuous color scales ----
con_color_plot



# Accessibility tip:

```


## `{paletteer}`

A few different ways to use `{paletteer}` palettes, but my preferred approach takes two steps: 

1. create vector of colors using `paletteer_*()`
2. apply your palette using the appropriate `ggplot::scale_*()` function


```{r}
#| eval: true
#| echo: true
# import {paletteer} ----
library(paletteer)
```


```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false


# Discrete Data ------

# Step 1: Create palette
pal_d 

#Step 2 Apply palette using `color` variant 
cat_color_plot 

#Step 2 Apply palette using `fill` variant)
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_histogram() 
```



```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

# Continuous Data ------

## Step 1:  create palette
pal_c 

#Step 2: Apply palette using `gradientn` variant
con_color_plot 

#Step 2: Apply palette using `stepsn` variant
con_color_plot 
```

## Some final palette tips

### Save palette outside of plot

```{r}
#| eval: true
#| echo: true
#| warning: false
# create palette ----
my_palette <- c("#32DE8A", "#E36414", "#0F4C5C")

# plot with custom palette ----
cat_color_plot 



```

### Set color names

```{r}
#| eval: true
#| echo: true
#| warning: false
# create palette ----
my_palette_named <- c("#32DE8A", "#E36414", "#0F4C5C")

# apply to plot (all penguins) ----
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_palette_named)

# apply to plot (just adelie & gentoo) ----
penguins |> 
  filter(species != "Chinstrap") |> 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_palette_named)
```

